Correctly Optimizing Abstract Abstract Machines

### 2013-04-02 paper/paper.tex
More work on initial sections.
Conflicts:

	paper/paper.tex

\authorinfo{J. Ian Johnson}
           {Northeastern University}
           {ianj@ccs.neu.edu}
\authorinfo{Nicholas Labich}
           {Northeastern University}
           {labichn@ccs.neu.edu}
\authorinfo{Matthew Might}
           {University of Utah}
           {might@cs.utah.edu}
\authorinfo{David Van Horn}
           {Northeastern University}
           {dvanhorn@ccs.neu.edu}
\maketitle

@@

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr,\mkaddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
\end{align*}
These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
used to determinize the choice of fresh addresses. The $\sqcup$ on
stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
relation is non-deterministic in its choice of addresses, however it
must always choose a fresh address when allocating a continuation or
variable binding.  If we consider machine states equivalent up to
consistent renaming and fix an allocation scheme, this relation
defines a deterministic machine (the relation is really a function).

@@

{\small
\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but one
that is too inefficient to run, even on small examples.  Through this
section, we explain a succession of approximations to reach a more
appropriate baseline analysis.
%
We ground this path by first formulating the analysis in terms of a
classic fixed-point computation.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
% \begin{equation*}
 $\eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}$
%\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

@@

In the next subsection, we fix this with store widening to reach polynomial
(albeit of high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it installs this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
{\small
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}}
%
The new store is computed as the least upper bound of all
stores after stepping in order to have only one, and keep the
semantics sound.

@@

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to contain this address
to remember where to retreive values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is demanded in order to continue with
evaluation. More generally, this is because applied functions are in
\emph{strict position} (more about that later).

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics;
 \item lazy non-determinism; and
 \item abstract compilation.
\end{enumerate}

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic worklist
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

The state space changes from a store and set of contexts to a set
of seen abstract states (context plus store), $S$, a set of contexts to
step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
\end{equation*}

We constantly see more states, so $S$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
%
\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}

The next store is the least upper bound of all the stores in $I$:

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  S' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\subsection{Imperative, pre-allocated data structures}

>>>>

\authorinfo{J. Ian Johnson \and Nicholas Labich \and Matthew Might \and David Van Horn}
           {}
           {}
%% \authorinfo{J. Ian Johnson}
%%            {Northeastern University}
%%            {ianj@ccs.neu.edu}
%% \authorinfo{Nicholas Labich}
%%            {Northeastern University}
%%            {labichn@ccs.neu.edu}
%% \authorinfo{Matthew Might}
%%            {University of Utah}
%%            {might@cs.utah.edu}
%% \authorinfo{David Van Horn}
%%            {Northeastern University}
%%            {dvanhorn@ccs.neu.edu}
\maketitle

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

@@

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie, $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but not an
efficient one.  Through
this section, we explain a succession of approximations
to reach the baseline analysis.  At each stage, we measure performance to
identify the importance of each choice.
%
We ground this path by first formulating the analysis in terms of a classic
fixed-point computation.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

@@

In the next subsection, we fix this with store widening and reach a polynomial
(albeit of high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to a merely
daunting polynomial.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it will install this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \widehat{\State} \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to store this address
to remember where to retreive values from in the store, though some allocation
strategies allow this representation to be ellided.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
%\newcommand{\ext}[3]{ext(#1,#2,#3)}

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\\
\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
We associate the value in hand with the expression that produced
it. Also, we fan out to all functions that are associated with the
original function position, because applied functions are in
\emph{strict position} (more about that later).

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, preallocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic ``worklist''
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
              S' &= \setof{(c,\msto') \mid c \in F'} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\subsection{Imperative, preallocated data structures}
